The \dealii Library, Version 9.4

### 2022-06-27 9.4/paper.tex
A paper reference should be managed with \cite
Last read.

While all of these major changes are discussed in detail in
Section~\ref{sec:major}, there
are a number of other noteworthy changes in the current \dealii release,
which we briefly outline in the remainder of this section:
%
\begin{itemize}
%  \item \texttt{AffineConstraints::make\_consistent\_in\_parallel()} allows
%  to make constraints consistent in parallel.
  \item The \texttt{DataOutResample} class interpolates values defined on one
  triangulation onto a second potentially unrelated triangulation.
  By using this class, one can output the result obtained on an
  unstructured mesh on a structured one (which might facilitate a more
  memory-efficient storage format, for example, if this second
  triangulation is a uniformly refined rectangle or box), or one can create a slice in 3D.
  \item The new member function \texttt{find\_point\_owner\_rank()} of
        \texttt{parallel\allowbreak ::distributed::\allowbreak Tri\allowbreak angulation} allows one to find the MPI
  ranks of cells containing specified points.
  It is communication-free and leverages the functionality of \pfrst (>v.2.2).
  Its algorithm is described in \cite{burstedde2020parallel}. This information will
  enable efficient construction of the
  communication pattern used in the class \texttt{Utilities::\allowbreak MPI::\allowbreak RemotePointEvaluation}. Furthermore, this function could be used in the future to allow
  particle simulations in which particle movement is not
  limited by CFL conditions, as done in \cite{mirzadeh2016parallel}.
  \item The new function
  \texttt{GridGenerator::pipe\_junction()}
  generates a triangulation of three cone-shaped pipes that cross at a bifurcation point in any possible configuration.
  A manifold description is applied to the boundary, which can be extended into the volume via transfinite interpolation \cite{Gordon82} using the \texttt{TransfiniteInterpolationManifold} class \cite{dealII90}.
  \item A new DoF renumbering function \texttt{DoFRenumbering::support\_point\_wise()} which groups together
  shape functions by their support point. This functionality is useful in both developing nodal schemes since, e.g., the $x$, $y$, and $z$
  components at a point will be consecutive in the solution vector. It also improves interoperability with external libraries which expect
  data in this format.
  \item The \texttt{FEInterfaceValues} class, which computes common quantities at the interface of two cells, has been overhauled to make it
  more consistent with the rest of the library and use more intuitive names for functions. For example,
  \texttt{FEInterfaceValues::jump\_gradient()} is now \texttt{FEInterfaceValues::jump\_in\_shape\_gradients()}. Several new
  functions, such as \texttt{FEInterfaceValues::get\_jump\_in\_function\_values()}, have also been added.
  \item The \texttt{MeshWorker::ScratchData} and \texttt{MeshWorker::CopyData} have been made $hp$-compatible,
  and support face integration where the integration rule and mapping differs on either side of an interface.
  The \texttt{MeshWorker::CopyData} class has also been made compatible with complex numbers.
  \item Vectors attached to \texttt{DataOut} do not need to be in ghosted state anymore. Internally, we create a copy of the vector with appropriate ghosting.
\end{itemize}
%
The changelog lists more than 100 other features and bugfixes.

@@

\begin{itemize}
\item Experimental support for locally refined meshes: For finite
  element computations on locally refined
meshes, one needs (i) the possibility to locally refine the mesh (see
Figure~\ref{fig:refinement}), and (ii) appropriate hanging-node
constraints. Both are now available in 2D; for 3D, the implementation
of the constraint definitions is still in progress.
\item \texttt{QIteratedSimplex} allows to build composite simplex quadrature rules.
\item The new wrappers to the \texttt{CGAL} library allow to simply create simplex
meshes. For more details, see Section~\ref{sec:cgalwrappers}.
\end{itemize}

@@

There are now new utility functions in the
\texttt{RepartitioningPolicyTools} namespace that can be
used to create a new \texttt{p::f::T} instance,
given a distributed triangulation (\texttt{p::f::T} or
\texttt{p::d::T}) and a vector that indicates the designated owner processes of locally
owned cells. This approach allows for partitioning \texttt{p::f::T}
based on arbitrary criteria. The workflow is shown in the following listing:
\begin{c++}
// original triangulation:
parallel::distributed::Triangulation<dim> tria(communicator);

// select a partitioning policy and use it:
const RepartitioningPolicyTools::CellWeightPolicy<dim> policy(tria, fu);
const auto construction_data = TriangulationDescription::Utilities::
  create_description_from_triangulation(tria, policy.partition(tria));

// create the new triangulation:
parallel::fullydistributed::Triangulation<dim> tria_pft(comm);
tria_pft.create_triangulation(construction_data);
\end{c++}

@@

The matrix-free infrastructure of \dealii is used to solve problems
without assembling matrices, providing only the action of the matrix
instead. It has seen a number of new features in this release, the most
notable ones being:
\begin{itemize}
\item Improved support for computations with Hessians of shape functions: just like values and gradients, Hessians can be
now evaluated and integrated during matrix-free loops
both for cells and faces. This could enable, for example, writing
a matrix-free version of step-47, which solves the biharmonic equation with the discontinuous Galerkin method.
\item Cell-centric loops now also allow access to gradients and Hessians
of neighboring cells on faces. The major difficulty here was the relative orientation
of cells on unstructured meshes.
\item Users can now create their own cell batches, by providing \texttt{FEEvaluation::reinit()} a list of cell IDs. \texttt{FEEvaluation}
accesses the appropriate data and reshuffles mapping data accordingly on
the fly in order to enable vectorization over cells. The new feature is useful in several
contexts, for example for sharp interfaces (e.g., two-phase flow
or shock capturing), where one needs to treat cells that are ``cut'' by
the interface in a special way. A challenge is that cell batches
might contain cut or non-cut cells, effectively counteracting the cross-cell vectorization.
Previous functionality has provided the option of masking certain cells in cell batches, which works well if
the code paths do not diverge too much, or to
categorize cells during \texttt{MatrixFree::reinit()} in such a way that mixed cell batches do not occur. However, \texttt{MatrixFree::reinit()}
might be too expensive if recategorization needs to happen very frequently to follow the dynamics of a system, e.g., in each time step. Despite some overhead compared to static matrix-free loops, the new feature can be the best option in such dynamic scenarios.
\item Initial support for H(div)-conforming elements with Piola transform, based on Raviart--Thomas finite element spaces with the updated class \texttt{FE\_RaviartThomasNodal}, has been added. This feature is currently limited to meshes in standard orientation and affine geometries. Full support and performance optimizations will be provided in a future release.
\item Selected matrix-free algorithms can now exploit additional data locality
  between the matrix-vector product and vector operations happening nearby in
  an algorithm. Interfaces have been added to both the
  \texttt{PreconditionChebyshev} class (a frequently used smoother in multigrid
  methods) and the conjugate gradient implementation in \texttt{SolverCG}.
  From the user's perspective, an operator needs to define a \texttt{vmult} operation,
  taking two additional \texttt{std::function} arguments. The first function
  defines the operation to be scheduled on the vector entries before the
  matrix-vector product touches them, and the second what happens
  afterward. The new features also include a renumbering to maximize data
  locality. The theory is described in the
  contribution~\cite{kronbichler2022cg}.
\end{itemize}
Besides these new features, we improved the performance of
hanging-node-constraint evaluation on the CPU. Instead of performing
quasi-dense matrix-vector multiplications~\cite{KronbichlerKormann2012}, we now use an
approach based on in-place interpolation and sum factorization, similar
to what was already done in the GPU code~\cite{ljungkvist2017matrix}. In \cite{munch2022hn}, the algorithm is described
and performance numbers are shown, indicating a reduction
of overhead of cells with hanging nodes by a factor of ten.

Finally, we have performed a major restructuring of the internals
of the \texttt{FEEvaluation} classes. This reduces some overheads for low polynomial degrees and will enable us to add support for new element types in the future.

@@

In \cite{munch2022gc}, the performance of the local-smoothing and global-coarsening
infrastructure of \dealii was compared for locally refined meshes. The results indicate that
the local definition of multigrid levels might introduce load imbalances
in the case of local smoothing so that global coarsening is favorable despite
potentially more expensive intergrid transfers. In order to judge the benefits
of one approach against the other, \dealii provides new functions
\texttt{workload\_imbalance()} and \texttt{vertical\_communication\_efficiency()}
in the \texttt{MGTools} namespace for the  estimation of the imbalance during, e.g.,
smoothing or the
communication efficiency during intergrid transfer, purely based on the given mesh.
% MK: I would not add this part, it does not really fit into this paper as
% there is nothing to report at this point (and we should then add references
% to the actual literature).
%The publication \cite{munch2022gc} also points out that not
%all types of smoothers are applicable for global coarsening due to the
%presence of hanging nodes, which is a motivation to add new smoother types
%to \dealii in the future.

@@

Several classes have been added to the \texttt{NonMatching} namespace to enable the use of cut finite element methods~\cite{burman_cutfem_2015}.
In the literature, these types of methods are also referred to as immersed, extended, or fictitious finite element methods.
Here, the domain, $\Omega$, is immersed in the background mesh, as illustrated in Fig.~\ref{fig:immersed-domain}.
Often, one solves for the degrees of freedom of the smallest submesh
which completely covers the domain, i.e., the blue and yellow cells of
Fig.~\ref{fig:location-to-level-set}.
The bilinear form in the weak form would then, for example, look like
\begin{equation}
  a(u,v) = (\nabla u, \nabla v)_\Omega - (\partial_n u, v)_\Gamma + \ldots
\end{equation}
Thus, when assembling on a cut cell, we are
required to integrate over the part of the domain and the part of the boundary, $\Gamma = \partial \Omega$, that falls inside the cell:
$K\cap \Omega$ and $K \cap \Gamma$.
Many of the new classes that support these operations assume that the domain is described by a level set function,
$\psi : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}\label{eq:levelset}
  \Omega = \{x \in \mathbb{R}^d : \psi(x)<0\},
  \qquad
  \Gamma = \{x \in \mathbb{R}^d : \psi(x) = 0\}.
\end{align}

@@

The Computational Geometry Algorithms Library (CGAL, \url{https://www.cgal.org/}) is a widely used
library to describe geometries and meshes \cite{cgal-user-ref}. \dealii now
has wrappers for CGAL classes and functions, provided in the new
namespace \texttt{CGALWrappers}: they implement functionality
spanning from mesh generation to boolean operations between
triangulations and cells. These wrappers are enabled only if \dealii
is compiled with \texttt{C++17}. \textit{Note: This feature is still experimental and
interfaces might change during the next release cycle.}

The main mesh generation function is
\texttt{GridGenerator::implicit\_function()}, which creates a \texttt{Triangulation<dim,3>} out of the zero level set of an implicit function $\psi$ similar to \eqref{eq:levelset}.
For \texttt{dim==3}, the mesh consists of tetrahedra. A prototypical use case is the following, where the surface is the zero level set of Taubin's heart function $f=\bigl ( x^2 + \frac{9y^2}{4} +z^2 -1 \bigr ) -x^2 z^3 - \frac{9y^2z^3}{80}$. The resulting \texttt{Triangulation<3>} is shown in Figure~\ref{fig:heart_tria}
and the required steps are:
\begin{c++}
// 1) An implicit function, e.g., Taubin's heart surface (not shown)
ImplicitFunction implicit_fu;

// 2) configure output mesh (optional)
CGALWrappers::AdditionalData<3> data; data.cell_size = .05;

// 3) create mesh
Triangulation<3> tria;
GridGenerator::implicit_function(tria, implicit_fu, data, {0, 0, 0}, 10.0);
\end{c++}

@@

CGAL also provides Boolean operations on meshes: they can be accessed via 
the utility function
\texttt{CGALWrappers::\allowbreak{}compute\_boolean\_operation()}. The
available operations are \textit{co-refinement}, \textit{intersection},
\textit{union}, and \textit{difference}. Oftentimes, Boolean
operations and co-refinement around the intersection produces
badly shaped mesh cells. To overcome this issue, one can use \texttt{CGALWrappers::\allowbreak{}remesh\_surface()}.
A possible workflow to create a good-quality mesh of the union of a cube and
a sphere mesh is given in the following listing:
\begin{c++}
// 1) create deal.II triangulations, e.g., cube and sphere (not shown)  
Triangulation<spacedim> tria0, tria1;

// 2) convert to CGAL surface meshes (assuming Kernel is already defined) 
CGAL::Surface_mesh<Kernel> surface_mesh0, surface_mesh1;
CGALWrappers::dealii_tria_to_cgal_surface_mesh(tria0, surface_mesh0);
CGALWrappers::dealii_tria_to_cgal_surface_mesh(tria1, surface_mesh1);

// 3) compute the union of the two meshes
CGALWrappers::compute_boolean_operation(surface_mesh0, surface_mesh1,
  BooleanOperation::compute_union, out_mesh);

// 4) convert CGAL surface mesh to deal.II surface mesh
Triangulation<2, 3> tria_out;
CGALWrappers::cgal_surface_mesh_to_dealii_triangulation(out_mesh, tria_out);

// 5) convert surface to volume mesh via surface_mesh_to_volumetric_mesh()
\end{c++}
The output of the Boolean operation is shown in Fig.~\ref{fig:corefinement};
Fig.~\ref{fig:corefinement_remeshed} shows the mesh after remeshing.
\begin{figure}
  \centering
  \begin{subfigure}[b]{0.28\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/heart_implicit.png}
    \caption{\label{fig:heart_tria}}
  \end{subfigure}\qquad
  \hfill
  \begin{subfigure}[b]{0.28\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/intersection_cube_sphere_mesh.png}
    \caption{\label{fig:corefinement}}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.35\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/cube_sphere_remeshed.png}
    \caption{ \label{fig:corefinement_remeshed}}
  \end{subfigure}
  \caption{\it (a) Triangulation created by filling a heart-shaped surface implicitly described by a function $f$. (b) Union of a cube with a sphere with badly shaped cells at the intersection. (c) Remeshed version of the same triangulation.}
\end{figure}

The function \texttt{CGALWrappers::compute\_quadrature\_on\_boolean\_operation()} returns a quadrature that allows exact integration of polynomials on polyhedral elements created by a \texttt{Boolean\allowbreak Operation} between \dealii cells.
The quadrature rule is built by 1) subdividing the polyhedral region in tetrahedra, 2) computing on each a \texttt{QGaussSimplex<3>} quadrature rule via \texttt{QSimplex<3>::\allowbreak{}compute\_affine\_transformation()}, and 3)
collecting all in a single \texttt{Quadrature<3>} object on the \emph{physical} cell.

@@

In the release described herein, we have reorganized the data
structures and optimized the algorithms that support using particles
in \dealii. In our previously reported improvements~\citep{dealII93} we stored all particle data as a separate contiguous array for each particle property, but particle identifiers (IDs) were still stored in a multimap tree-like structure.

@@

\begin{itemize}
\item All particle data (both their identifiers and their actual data) are now stored as separate and contiguous arrays in memory, which improves spatial locality for better prefetching of data and makes iterating over particles extremely efficient.
\item The choice of a list container that only includes entries for cells that contain particles means iteration is efficient, even if many cells in the domain do not contain any particles (as can be the case for discrete element methods~\cite{golshan2022lethe}).
\item Creating separate arrays for each cell allows us to easily move particle IDs from one cell to another as a local operation, affecting only the two cell containers in question. We take care to reuse allocated memory to minimize the number of memory reallocations.
\item The separate cache structure that contains entries for each cell allows quick random access to the particles of a particular cell, and also allows to quickly determine if a particular cell has particles at all.
\end{itemize}

@@

   Sending large messages over MPI or reading and writing large buffers using MPI I/O,
   especially when dealing with small datatypes like \texttt{MPI\_CHAR},
   can often require processing more than $2^{31}$ objects at once. Prior
   to the recent MPI-4 standard, which introduced a new set of \texttt{MPI\_*\_c()} functions, all MPI functions expected the ``count''
   to be a signed integer. This limits the number of objects to $2^{31}$,
   i.e., writing of up to 2 GB at once if the datatype is \texttt{MPI\_CHAR}.

   Several places in the library could run into the issue mentioned above,
   for example when producing large graphical output (per rank), large
   checkpointing (per rank), or when broadcasting large datasets for lookup tables.

   With this release we introduce a new module \texttt{Utilities::MPI::LargeCount} which enables sending and receiving MPI messages and I/O containing more than $2^{31}$ objects by implementing
   the before-mentioned \texttt{MPI\_*\_c()} functions, e.g., \texttt{MPI\_Bcast\_c()}, using MPI-3 features if necessary. The
   solution is based on custom datatypes as described in~\ref{hammond2014int_max}.

   This functionality is now used in all places in the library, where
   large counts might be necessary (\texttt{DataOut::write\_vtu\_in\_parallel()} for graphical output, checkpointing of \texttt{Triangulation} objects, etc.).
   %
   While implementing these changes we also worked on the following:
   \begin{itemize}
    \item Testing of large I/O with large chunks per MPI rank (>2 GB) and
    large total sizes (4 GB +). Several instances of 32-bit data types for
    offsets were changed to 64-bit to correctly support files larger
    than 4 GB (HDF5 output, VTU output, checkpointing).
    \item Performance testing of MPI I/O routines used for parallel VTU output with large performance improvements by switching from a shared
    file pointer to individual file pointers, see Table~\ref{tab:mpi-io}.
   \end{itemize}

\begin{table}
  \caption{\it Performance comparison of writing large VTU graphical output using MPI I/O on
  TACC Frontera /scratch1/ with 16 file servers (``OSTs'') and a theoretical peak performance of 50 GB/s. The ROMIO version used enforces sequential writes when using \texttt{MPI\_File\_write\_ordered()}.}
  \label{tab:mpi-io}

>>>>

While all of these major changes are discussed in detail in
Section~\ref{sec:major}, there
are a number of other noteworthy changes in the current \dealii release,
which we briefly outline in the remainder of this section:
%
\begin{itemize}
%  \item \texttt{AffineConstraints::make\_consistent\_in\_parallel()} allows
%  to make constraints consistent in parallel.
  \item The \texttt{DataOutResample} class interpolates values defined on one
  triangulation onto a second, potentially unrelated triangulation.
  By using this class, one can output the result obtained on an
  unstructured mesh on a structured one (which might facilitate a more
  memory-efficient storage format, for example, if this second
  triangulation is a uniformly refined rectangle or box), or one can create a slice in 3D.

\item The new member function \texttt{find\_point\_owner\_rank()} of
        \texttt{parallel\allowbreak ::distributed::\allowbreak Tri-\allowbreak angulation} allows one to find the MPI
  ranks of cells containing specified points.
  It is communication-free and leverages the functionality of \pfrst (>v.2.2).
  Its algorithm is described in \cite{burstedde2020parallel}. This information will
  enable efficient construction of the
  communication pattern used in the class \texttt{Utilities::\allowbreak MPI::\allowbreak RemotePointEvaluation}. Furthermore, this function could be used in the future to allow
  particle simulations in which particle movement is not
  limited by CFL conditions, as done in \cite{mirzadeh2016parallel}.

\item The new function
  \texttt{GridGenerator::pipe\_junction()}
  generates a triangulation of three cone-shaped pipes that cross at a bifurcation point in any possible configuration.
  A manifold description is applied to the boundary, which can be extended into the volume via transfinite interpolation \cite{Gordon82} using the \texttt{TransfiniteInterpolationManifold} class \cite{dealII90}.
  \item A new DoF renumbering function \texttt{DoFRenumbering::support\_point\_wise()} which groups together
  shape functions by their support point. This functionality is useful in both developing nodal schemes since, e.g., the $x$, $y$, and $z$
  components at a point will be consecutive in the solution vector. It
  also improves interoperability with those external libraries that expect
  data in this format.
  \item The \texttt{FEInterfaceValues} class, which computes common quantities at the interface of two cells, has been overhauled to make it
  more consistent with the rest of the library and use more intuitive names for functions. For example,
  \texttt{FEInterfaceValues::jump\_gradient()} is now \texttt{FEInterfaceValues::jump\_in\_shape\_gradients()}. Several new
  functions, such as \texttt{FEInterfaceValues::get\_jump\_in\_function\_values()}, have also been added.
  \item The \texttt{MeshWorker::ScratchData} and \texttt{MeshWorker::CopyData} have been made $hp$-compatible,
  and support face integration where the integration rule and mapping differs on either side of an interface.
  The \texttt{MeshWorker::CopyData} class has also been made compatible with complex numbers.
  \item Vectors attached to \texttt{DataOut} do not need to be in
    ghosted state anymore. Internally, we create a copy of the vector
    with appropriate ghost elements.
\end{itemize}
%
The changelog lists more than 100 other features and bugfixes.

@@

\begin{itemize}
\item Experimental support for locally refined meshes: For finite
  element computations on locally refined
meshes, one needs (i) the possibility to locally refine the mesh (see
Figure~\ref{fig:refinement}), and (ii) appropriate hanging-node
constraints. Both are now available in 2D; for 3D, the implementation
of the constraint definitions is still in progress.
\item \texttt{QIteratedSimplex} allows to build composite simplex quadrature rules.
\item The new wrappers to the \texttt{CGAL} library allow for the easy
  creation of simplex
meshes. For more details, see Section~\ref{sec:cgalwrappers}.
\end{itemize}

@@

There are now new utility functions in the
\texttt{RepartitioningPolicyTools} namespace that can be
used to create a new \texttt{p::f::T} instance,
given a distributed triangulation (\texttt{p::f::T} or
\texttt{p::d::T}) and a vector that indicates the designated owner processes of locally
owned cells. This approach allows for partitioning \texttt{p::f::T}
based on arbitrary criteria. The workflow is shown in the following listing:
\begin{c++}
// Original triangulation:
parallel::distributed::Triangulation<dim> tria(communicator);

// Select a partitioning policy and use it:
const RepartitioningPolicyTools::CellWeightPolicy<dim> policy(tria, fu);
const auto construction_data = TriangulationDescription::Utilities::
  create_description_from_triangulation(tria, policy.partition(tria));

// Create the new triangulation:
parallel::fullydistributed::Triangulation<dim> tria_pft(comm);
tria_pft.create_triangulation(construction_data);
\end{c++}

@@

The matrix-free infrastructure of \dealii is used to solve problems
without assembling matrices, providing only the \textit{action} of the matrix
instead. It has seen a number of new features in this release, the most
notable ones being:
\begin{itemize}
\item Improved support for computations with Hessians of shape functions: Just like values and gradients, Hessians can be
now evaluated and integrated during matrix-free loops
both for cells and faces. This could enable, for example, writing
a matrix-free version of step-47, which solves the biharmonic equation with the discontinuous Galerkin method.
\item Cell-centric loops now also allow access to gradients and Hessians
of neighboring cells on faces. The major difficulty here was the relative orientation
of cells on unstructured meshes.
\item Users can now create their own cell batches, by providing \texttt{FEEvaluation::reinit()} a list of cell IDs. \texttt{FEEvaluation}
accesses the appropriate data and reshuffles mapping data accordingly on
the fly in order to enable vectorization over cells. The new feature is useful in several
contexts, for example for sharp interfaces (e.g., two-phase flow
or shock capturing), where one needs to treat cells that are ``cut'' by
the interface in a special way. A challenge is that cell batches
might contain cut or non-cut cells, effectively counteracting the cross-cell vectorization.
Previous functionality has provided the option of masking certain cells in cell batches, which works well if
the code paths do not diverge too much, or to
categorize cells during \texttt{MatrixFree::reinit()} in such a way that mixed cell batches do not occur. However, \texttt{MatrixFree::reinit()}
might be too expensive if recategorization needs to happen very frequently to follow the dynamics of a system, e.g., in each time step. Despite some overhead compared to static matrix-free loops, the new feature can be the best option in such dynamic scenarios.
\item Initial support for H(div)-conforming elements with Piola transform, based on Raviart--Thomas finite element spaces with the updated class \texttt{FE\_RaviartThomasNodal}, has been added. This feature is currently limited to meshes in standard orientation and affine geometries. Full support and performance optimizations will be provided in a future release.
\item Selected matrix-free algorithms can now exploit additional data locality
  between the matrix-vector product and vector operations happening nearby in
  an algorithm. Interfaces have been added to both the
  \texttt{PreconditionChebyshev} class (a frequently used smoother in multigrid
  methods) and the conjugate gradient implementation in \texttt{SolverCG}.
  From the user's perspective, an operator needs to define a \texttt{vmult} operation,
  taking two additional \texttt{std::function} arguments. The first function
  defines the operation to be scheduled on the vector entries before the
  matrix-vector product touches them, and the second what happens
  afterward. The new features also include a renumbering to maximize data
  locality. The theory is described in the
  contribution~\cite{kronbichler2022cg}.
\end{itemize}
Besides these new features, we improved the performance of
hanging-node-constraint evaluation on the CPU. Instead of performing
quasi-dense matrix-vector multiplications~\cite{KronbichlerKormann2012}, we now use an
approach based on in-place interpolation and sum factorization, similar
to what was already done in the GPU code~\cite{ljungkvist2017matrix}. In \cite{munch2022hn}, the algorithm is described
and performance numbers are shown, indicating a reduction
of overhead of cells with hanging nodes by a factor of ten.

Finally, we have performed a major restructuring of the internals
of the \texttt{FEEvaluation} classes. This reduces some overhead for low polynomial degrees and will enable us to add support for new element types in the future.

@@

In \cite{munch2022gc}, the performance of the local-smoothing and global-coarsening
infrastructure of \dealii was compared for locally refined meshes. The results indicate that
the local definition of multigrid levels might introduce load imbalances
in the case of local smoothing so that global coarsening is favorable despite
potentially more expensive intergrid transfers. In order to assess the benefits
of one approach against the other, \dealii provides new functions
\texttt{workload\_imbalance()} and \texttt{vertical\_communication\_efficiency()}
in the \texttt{MGTools} namespace for the  estimation of the imbalance during, e.g.,
smoothing or the
communication efficiency during intergrid transfer, purely based on the given mesh.
% MK: I would not add this part, it does not really fit into this paper as
% there is nothing to report at this point (and we should then add references
% to the actual literature).
%The publication \cite{munch2022gc} also points out that not
%all types of smoothers are applicable for global coarsening due to the
%presence of hanging nodes, which is a motivation to add new smoother types
%to \dealii in the future.

@@

Several classes have been added to the \texttt{NonMatching} namespace to enable the use of cut finite element methods~\cite{burman_cutfem_2015}.
In the literature, these types of methods are also referred to as immersed, extended, or fictitious finite element methods.
Here, the domain, $\Omega$, is immersed in the background mesh, as illustrated in Fig.~\ref{fig:immersed-domain}.
Often, one solves for the degrees of freedom of the smallest submesh
which completely covers the domain, i.e., the blue and yellow cells of
Fig.~\ref{fig:location-to-level-set}.
The bilinear form in the weak form would then, for example, look like
\begin{equation}
  a(u,v) = (\nabla u, \nabla v)_\Omega - (\partial_n u, v)_\Gamma + \ldots
\end{equation}
Thus, when assembling on a cut cell $K$, we are
required to integrate over the part of the domain and the part of the boundary, $\Gamma = \partial \Omega$, that falls inside the cell:
$K\cap \Omega$ and $K \cap \Gamma$.
Many of the new classes that support these operations assume that the domain is described by a level set function,
$\psi : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}\label{eq:levelset}
  \Omega = \{x \in \mathbb{R}^d : \psi(x)<0\},
  \qquad
  \Gamma = \{x \in \mathbb{R}^d : \psi(x) = 0\}.
\end{align}

@@

The Computational Geometry Algorithms Library (CGAL, \url{https://www.cgal.org/}) is a widely used
library to describe geometries and meshes \cite{cgal-user-ref}. \dealii now
has wrappers for CGAL classes and functions, provided in the new
namespace \texttt{CGALWrappers}: they implement functionality
spanning from mesh generation to boolean operations between
triangulations and cells. These wrappers are enabled only if \dealii
is compiled with \texttt{C++17}. \textit{This feature is still experimental and
interfaces might change during the next release cycle.}

The main mesh generation function is
\texttt{GridGenerator::implicit\_function()}, which creates a \texttt{Triangulation<dim,3>} out of the zero level set of an implicit function $\psi$ similar to \eqref{eq:levelset}.
For \texttt{dim==3}, the mesh consists of tetrahedra. A prototypical use case is the following, where the surface is the zero level set of Taubin's heart function $f=\bigl ( x^2 + \frac{9y^2}{4} +z^2 -1 \bigr ) -x^2 z^3 - \frac{9y^2z^3}{80}$. The resulting \texttt{Triangulation<3>} is shown in Figure~\ref{fig:heart_tria}
and the required steps are:
\begin{c++}
// 1) An implicit function, e.g., Taubin's heart surface (not shown)
ImplicitFunction f;

// 2) Configure output mesh (optional)
CGALWrappers::AdditionalData<3> data; data.cell_size = .05;

// 3) Create mesh
Triangulation<3> tria;
GridGenerator::implicit_function(tria, f, data, /* more arguments */...);
\end{c++}

@@

CGAL also provides Boolean operations on meshes: they can be accessed via 
the utility function
\texttt{CGALWrappers::\allowbreak{}compute\_boolean\_operation()}. The
available operations are \textit{co-refinement}, \textit{intersection},
\textit{union}, and \textit{difference}. Oftentimes, Boolean
operations and co-refinement around the intersection produces
badly shaped mesh cells. To overcome this issue, one can use \texttt{CGALWrappers::\allowbreak{}remesh\_surface()}.
A possible workflow to create a high-quality mesh of the union of a cube and
a sphere mesh is given in the following listing:
\begin{c++}
// 1) Create deal.II triangulations, e.g., cube and sphere (not shown)  
Triangulation<spacedim> tria0, tria1;

// 2) Convert to CGAL surface meshes (assuming Kernel is already defined) 
CGAL::Surface_mesh<Kernel> surface_mesh0, surface_mesh1;
CGALWrappers::dealii_tria_to_cgal_surface_mesh(tria0, surface_mesh0);
CGALWrappers::dealii_tria_to_cgal_surface_mesh(tria1, surface_mesh1);

// 3) Compute the union of the two meshes
CGALWrappers::compute_boolean_operation(surface_mesh0, surface_mesh1,
  BooleanOperation::compute_union, out_mesh);

// 4) Convert CGAL surface mesh to deal.II surface mesh
Triangulation<2, 3> tria_out;
CGALWrappers::cgal_surface_mesh_to_dealii_triangulation(out_mesh, tria_out);

// 5) Convert surface to volume mesh via surface_mesh_to_volumetric_mesh()
\end{c++}
The output of the Boolean operation is shown in Fig.~\ref{fig:corefinement};
Fig.~\ref{fig:corefinement_remeshed} shows the mesh after remeshing.
\begin{figure}
  \centering
  \begin{subfigure}[b]{0.28\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/heart_implicit.png}
    \caption{\label{fig:heart_tria}}
  \end{subfigure}\qquad
  \hfill
  \begin{subfigure}[b]{0.28\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/intersection_cube_sphere_mesh.png}
    \caption{\label{fig:corefinement}}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.35\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/cube_sphere_remeshed.png}
    \caption{ \label{fig:corefinement_remeshed}}
  \end{subfigure}
  \caption{\it (a) Triangulation created by filling a heart-shaped surface implicitly described by a function $f$. (b) Union of a cube with a sphere with badly shaped cells at the intersection. (c) Remeshed version of the same triangulation.}
\end{figure}

The function
\texttt{CGALWrappers::compute\_quadrature\_on\_boolean\_operation()}
returns a quadrature that allows exact integration of polynomials on
polyhedral elements created by a Boolean operation between \dealii cells.
The quadrature rule is built by (i) subdividing the polyhedral region in tetrahedra, (ii) computing on each a \texttt{QGaussSimplex<3>} quadrature rule via \texttt{QSimplex<3>::\allowbreak{}compute\_affine\_transformation()}, and (iii)
collecting all in a single quadrature object on the \emph{physical} cell.

@@

In the release described herein, we have reorganized the data
structures and optimized the algorithms that support using particles
in \dealii. In our previously reported implementation~\citep{dealII93} we stored all particle data as a separate contiguous array for each particle property, but particle identifiers (IDs) were still stored in a multimap tree-like structure.

@@

\begin{itemize}
\item All particle data (both their identifiers and their actual data) are now stored as separate and contiguous arrays in memory, which improves spatial locality for better prefetching of data and makes iterating over particles extremely efficient.
\item The choice of a list container that only includes entries for cells that contain particles means iteration is efficient, even if many cells in the domain do not contain any particles (as can be the case for discrete element methods~\cite{golshan2022lethe}).
\item Creating separate arrays for each cell allows us to easily move particle IDs from one cell to another as a local operation, affecting only the two cell containers in question. We take care to reuse allocated memory to minimize the number of memory reallocations.
\item The separate cache structure that contains entries for each cell allows quick random access to the particles of a particular cell, and also allows quickly determining if a particular cell has particles at all.
\end{itemize}

@@

   Sending large messages over MPI or reading and writing large buffers using MPI I/O,
   especially when dealing with small datatypes like \texttt{MPI\_CHAR},
   can often require processing more than $2^{31}$ objects at once. Prior
   to the recent MPI-4 standard, which introduced a new set of \texttt{MPI\_*\_c()} functions, all MPI functions expected the ``count''
   to be a signed integer. This limits the number of objects to $2^{31}$,
   i.e., writing of up to 2 GB at once if the datatype is \texttt{MPI\_CHAR}.
   Several places in the library could exceed this limit,
   for example when producing large graphical output, large
   checkpoint files, or when broadcasting large datasets such as lookup tables.

   With this release we introduce a new module \texttt{Utilities::MPI::LargeCount} which enables sending and receiving MPI messages and I/O containing more than $2^{31}$ objects by implementing
   the before-mentioned \texttt{MPI\_*\_c()} functions, e.g., \texttt{MPI\_Bcast\_c()}, using MPI-3 features if necessary. The
   solution is based on custom datatypes as described in~\cite{hammond2014int_max}.
This functionality is now used in all places in the library where
   large counts might be necessary (\texttt{DataOut::write\_vtu\_in\_parallel()} for graphical output, checkpointing of \texttt{Triangulation} objects, etc.).

   While implementing these changes we also worked on the following:
   \begin{itemize}
    \item Testing of large I/O with large chunks per MPI rank (>2 GB) and
    large total sizes (>4 GB). Several instances of 32-bit data types for
    offsets were changed to 64-bit to correctly support files larger
    than 4 GB (HDF5 output, VTU output, checkpointing).
    \item Performance testing of MPI I/O routines used for parallel VTU output with large performance improvements by switching from a shared
    file pointer to individual file pointers, see Table~\ref{tab:mpi-io}.
   \end{itemize}

\begin{table}
  \caption{\it Performance comparison of writing large VTU graphical output using MPI I/O on
  TACC Frontera \texttt{/scratch1} file system with 16 file servers (``OSTs'') and a theoretical peak performance of 50 GB/s. The ROMIO version used enforces sequential writes when using \texttt{MPI\_File\_write\_ordered()}.}
  \label{tab:mpi-io}
